Correctly Optimizing Abstract Abstract Machines

### 2013-06-13 paper/paper.tex
Improved visualization for baseline/lazy/lazy&compiled.

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

\item generate states faster.  We accomplish this by better algorithm
  design in the fixed-point computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
small example due to Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.
By generating significantly fewer states at a significantly faster
rate, we are able to achieve large performance improvements in terms
of both time and space.

@@

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains};
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its set of traces that arise from iterating the
machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,\footnote{
More generally, we can have stores map to any domain that forms
 a Galois connection with values, enabling $\interpdelta$ to
produce elaborate abstractions of base values (e.g., interval or octagon
abstractions). We use sets of values for a simpler exposition.
} not
  single values,

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it will install this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \widehat{\State} \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. For clarity, we will exhibit non-widened
semantics unless a technique specifically requires it. The final step
to the baseline takes the complexity from quartic to cubic in the
monovariant case.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to contain this address
to remember where to retreive values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
%\newcommand{\ext}[3]{ext(#1,#2,#3)}

HERE

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\mkaddr}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision.

\begin{align*}
\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic ``worklist''
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
              S' &= \setof{(c,\msto') \mid c \in F'} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would not otherwise happen, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. The semantics is different - we can't precisely answer
    questions about the traces of the reduction relation we defined.
    Our semantics has a drop-in replacement of timestamps for stores
    in the seen set ($S$), so we do not experience precision loss.}
\end{enumerate}

>>>>

\begin{figure}[t]
\small
\begin{center}
%\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of the rate of state
  transitions and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

@@

\item generate states faster.  We accomplish this by better algorithm
  design in the fixed-point computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the
small motivating example in Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.  By generating
significantly fewer states at a significantly faster rate, we are able
to achieve large performance improvements in terms of both time and
space.

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for Earl et. al. program.  Gray states
  follow variable references, {\tt ev} states are black, and all others
  are white. Part (a) shows the result of the baseline analyzer.  It
  has long ``corridor'' transitions and ``diamond'' subgraphs that
  fan-out from nondeterminism and fan-in from joins.  Part (b) shows
  the result of performing nondeterminism lazily and thus avoids many
  of the diamond subgraphs.  Part (c) shows the result of abstract
  compilation that removes interpretive overhead in the form of
  intermediate states, thus minimizing the corridor transitions.  The
  end result is a more compact abstraction of the program that can be
  generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev[^\mcntr]{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap[^\mcntr]{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Times} &\mcntr&\in&\Counter \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \wp(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains};
figure~\ref{fig:aam} defines the transition relation (unmentioned
components stay the same).  The evaluation of a program is defined as
its set of traces that arise from iterating the machine transition
relation. The $\traces$ function produces the set of all proofs of
reachability for any state $\mstate$ from the injection of program
$\mexp$ (from which one could extract a string of states). The machine
is a very slight variation on a standard abstract machine for ISWIM in
``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It can be
systematically derived from a definitional interpreter through a
continuation-passing style transformation and defunctionalization, or
from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,\footnote{
More generally, we can have stores map to any domain that forms
 a Galois connection with sets of values, enabling $\interpdelta$ to
produce elaborate abstractions of base values (e.g., interval or octagon
abstractions). We use sets of values for a simpler exposition.
} not
  single values,

@@

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
%%\co{\kmt,\mval,\msto} &\machstep& \ans{\msto,\mval}
%%\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

@@

{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\epsilon} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but the consequential accelerated convergence fits the M.O. of a widening operator.} during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it installs this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
{\small
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
%%\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}}
%
The new store is computed as the least upper bound of all
stores after stepping in order to have only one, and keep the
semantics sound.

@@

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to contain this address
to remember where to retrieve values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is necessary in order to continue with
evaluation.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics;
 \item lazy non-determinism; and
 \item abstract compilation.
\end{enumerate}

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic worklist
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $S$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
\end{equation*}

We constantly see more states, so $S$ is always growing. The frontier,
which is what remains to be done, changes. Let's start with the result
of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
%
\begin{equation*}
I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
\end{equation*}

The next store is the least upper bound of all the stores in $I$:

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  S' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. Our semantics has a drop-in replacement of timestamps for
    stores in the seen set ($S$), so we do not experience precision
    loss.}
\end{enumerate}
